---
title: "illnesses.qmd"
format: pdf
editor: visual
---

```{r}
library(tidyverse)
library(tidymodels)
```

```{r}
covid <- read_csv("Covid Data.csv")

```

```{r}
# 0 = did not die
# 1 = died
covid_health <- covid |>
  mutate(
    died = if_else(DATE_DIED == "9999-99-99", 0, 1)
  ) |>
  filter(PNEUMONIA != 97 & PNEUMONIA != 99 & PNEUMONIA != 98) |>
  filter(DIABETES != 97 & DIABETES != 99 & DIABETES != 98) |>
  filter(INMSUPR != 97 & INMSUPR != 99 & INMSUPR != 98) |>
  filter(HIPERTENSION != 97 & HIPERTENSION != 99 & HIPERTENSION != 98) |>
  filter(OTHER_DISEASE != 97 & OTHER_DISEASE != 99 & OTHER_DISEASE != 98) |>
  filter(CARDIOVASCULAR != 97 & CARDIOVASCULAR != 99 & CARDIOVASCULAR != 98) |>
  filter(RENAL_CHRONIC != 97 & RENAL_CHRONIC != 99 & RENAL_CHRONIC != 98) |>
  filter(ASTHMA != 97 & ASTHMA != 99 & ASTHMA != 98) |>
  filter(AGE != 97 & AGE != 99 & AGE != 98) |>
  filter(SEX != 97 & SEX != 99 & SEX != 98)

covid_health

```

```{r}
# pneumonia specifically influential bc it's a lung disease

covid_health |>
  mutate(PNEUMONIA = if_else(PNEUMONIA == 1, "Pneumonia", "No Pneumonia")) |>
  mutate(died = if_else(died == 1, "Died", "Alive")) |>
  ggplot(aes(x = PNEUMONIA, fill = died)) +
  geom_bar(position = "fill")  +
  labs(title = "Deaths by COVID-19 Based on Presence of Pneumonia", y = "Proportion of Patients that Died", x = "Presence of Pneumonia", fill = "Death from COVID-19")


```

```{r}
covid_long <- covid_health |>
  select(PNEUMONIA, DIABETES, ASTHMA, INMSUPR, HIPERTENSION, OTHER_DISEASE, CARDIOVASCULAR, RENAL_CHRONIC, died) |>
  pivot_longer(
    cols = c('PNEUMONIA', 'DIABETES', 'ASTHMA', 'INMSUPR', 'HIPERTENSION', 'OTHER_DISEASE', 'CARDIOVASCULAR', 'RENAL_CHRONIC'),
    names_to = "Health Condition",
    values_to = "Presence"
  ) |>
  mutate(Presence = if_else(Presence == 1, "Yes", "No"))

covid_long


```

```{r}

logit_mod_health <- glm(died ~ as.factor(PNEUMONIA) + as.factor(DIABETES) + as.factor(ASTHMA) + as.factor(INMSUPR) + as.factor(HIPERTENSION) + as.factor(CARDIOVASCULAR) + as.factor(RENAL_CHRONIC),
                        data = covid_health,
                        family = "binomial")

tidy(logit_mod_health)

```

```{r}
logit_health_aug <- augment(logit_mod_health)

logit_health_aug


```

```{r}
logit_health_aug <- logit_health_aug |>
  mutate(prob = exp(.fitted)/(1 + exp(.fitted)),
         pred_died = ifelse(prob > 0.5, "Died", "Did Not Die")) %>% 
  select(.fitted, prob, pred_died, died)

logit_health_aug


```

```{r}
table(logit_health_aug$pred_died, logit_health_aug$died)

```
##DEMOGRAPHICS:
```{r}
covid_health$PREGNANT <- NULL
covid_health
```
```{r}
covid_health<- covid_health |>
mutate(DIED = ifelse(DATE_DIED == '9999-99-99', 0, 1)) |>
mutate(SEX = as.factor(SEX)) |>
mutate(DIED = as.factor(DIED))
covid_health
```
```{r}
covid_health$SEX <- factor(covid_health$SEX, levels = c(1, 2), labels = c("Female", "Male"))
covid_health
```

```{r}
library(scales)
covid_health |>
filter(DIED == 1) |>
ggplot(aes(x = SEX, fill = SEX)) +
geom_bar(stat = "count", width = 0.6) + # Adjust bar width for aesthetics
scale_fill_manual(values = c("blue", "orange")) + # Change colors for clarity
labs(
title = "Number of Deaths by Sex",
x = "Gender",
y = "Count of Deaths"
) +
theme_minimal() +
theme(
text = element_text(size = 14), # Adjust text size for better readability
plot.title = element_text(hjust = 0.5), # Center the plot title
axis.title = element_text(size = 12), # Specify axis title size
legend.position = "none" # Remove legend if redundant
) +
scale_y_continuous(labels = comma)
```
```{r}
covid_summary <- covid_health |>
count(SEX, DIED) |>
group_by(SEX) |>
mutate(Proportion = n / sum(n))
# Plot
ggplot(covid_summary, aes(x = SEX, y = Proportion, fill = DIED)) +
geom_col() +
scale_y_continuous(labels = scales::percent_format()) +
labs(
title = "Proportion of COVID-19 Deaths by Sex",
subtitle = "Male Proportion of Deaths is Larger than Females",
x = "SEX",
y = "Proportion"
) +
scale_fill_manual(values = c("0" = "lightgrey", "1" = "red"),
labels = c("Survived", "Died")) +
theme_minimal() +
theme(legend.title = element_blank())
```

```{r}
covid_health |>
mutate(DIED = factor(DIED, levels = c(0, 1), labels = c("Didn't Die", "Died"))) |>
ggplot(aes(x = DIED, y = AGE, fill = DIED)) +
geom_boxplot(outlier.colour = "black", outlier.shape = 1) +
labs(title = "COVID-19 Death vs. Age Correlation",
subtitle = "Median age of the deceased patients is greater than that of the survived", x = "Death",
y = "Age") +
scale_fill_manual(values = c("Didn't Die" = "lightblue", "Died" = "pink")) +
theme_minimal() +
theme(legend.position = "none")
```
```{r}
ggplot(covid_health, aes(x = AGE, fill = DIED)) +
geom_histogram(binwidth = 5, position = "dodge") +
facet_wrap(~SEX) +
labs(title = "Age Distribution by Death Status and Gender",
x = "Age",
y = "Count") +
scale_fill_manual(values = c("0" = "lightblue", "1" = "red"))
```

```{r}
logit_mod_demo <- glm(DIED ~ as.factor(SEX) + AGE, data = covid_health, family = "binomial")
tidy(logit_mod_demo)
```
SEX male: Holding age constant, we predict the odds of a male patient in the COVID-19 dataset passing away to be around e^0.63753001(1.8918) times that of a female patient.
AGE: Holding sex constant, we predict that for each additional year in age of the patient, the
odds of passing away are multiplied by e^0.07605202(1.0790).
```{r}
logit_demo_aug <- augment(logit_mod_demo)
logit_demo_aug
```

```{r}
logit_demo_aug <- logit_demo_aug |>
mutate(prob = exp(.fitted)/(1 + exp(.fitted)),
pred_died = ifelse(prob > 0.5, "Died", "Did Not Die")) |>
select(.fitted, prob, pred_died, DIED)
logit_demo_aug
```
```{r}
logit_demo_aug |>
roc_auc(
truth = DIED,
prob, event_level = "second"
)
```

```{r}
logit_demo_aug |>
roc_curve(
truth = DIED,
.fitted,
event_level = "second"
) |>
autoplot()
```

The AUC I achieved is 0.8395. It means that there is approximately an 83.95% chance that the model will be able to distinguish between a patient who died and one who did not die from COVID-19.
```{r}
library(Stat2Data)
library(pROC)
emplogitplot1(DIED ~ AGE, data = covid_health, ngroups = 20,
main = "Linerity Satisfied Log(Odds) vs. AGE")
```
Since age is the only continous variable in the model, we must check our linearity assumption.
From the plot graphed above, we see that it displays a linear relationship between the log odds
of dying and age. The graph displays points that are relatively linearly uniform and follow the
trend of the line. From the graph, we do not see that it contain any distinct trend that would
challenge our assumption of linearity
